---
title: "Available Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Available Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r setup}
library(tidyverse)
library(dvwaste)
```

## Parameterised Bar Chart Function

This funtion is designed to filter and process data from the `coast_waste_10` dataset for specific countries, resulting in a tibble that can be used to create a bar chart. To use this feature, you must provide a single parameter in the form of a list or vector that contains the names of all selected countries. 

```
# usage
dvwaste::param_barchart(country_list = character(0))

```
Each element in the country list must match an entry in the 'Entity' column from the coast_waste_10 dataset. If you leave this parameter empty, data for all valid countries will be included by default. The output of this feature is a table with two columns: "Entity" (country names) and "Value" (mismanaged plastic waste in tonnes).

```
# A list of all available countries in the dataset
dvwaste::coast_waste_10$Entity

```
<details>
  <summary><i>See the full country list</i></summary>

```{r}
# A list of all available countries in the dataset
dvwaste::coast_waste_10$Entity
```
  </details>
#### Examples

```{r}

# select multiple countries
input_countries <- c("Bahamas", "Australia", "United States")
param_barchart(input_countries)

# Use default country_list value, return a table with all countries
param_barchart()
```

## Summary_Statistics_Function

This function summarizes either character/factor or numeric data, providing frequency, proportion, cumulative frequency, and other statistics based on the input type. It takes a numeric, character, or factor vector as argument. The funtion returns a data frame with statistics such as frequency, proportion, cumulative frequency, and other summary measures, depending on the input type.

When the input is a numerical vector, the function generates a table containing 5 summary descriptions, including the Mean, Median, SD, and N_Sample. On the other hand, if a character or factor vector is input, the function calculates the Frequency, Cumulative_Frequency, Proportion, and Cumulative_Proportion. The output for a character or vector input will be presented in a tidy format.

```
# usage
# @param x : A numeric, character, or factor vector.
dvwaste::descriptive_stats(x)

```

#### Examples

```{r}
library(tidyverse)
library(dvwaste)

# Example with a factor
factor_input <- factor(c("A", "B", "A", "C", "B"))
descriptive_stats(factor_input)

# Example with a character vector
character_input <- c("A", "B", "A", "C", "B")
descriptive_stats(character_input)

# Example with a numeric vector
numeric_input <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
descriptive_stats(numeric_input)
```

### Reference 

**Tidyverse**
Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester
  J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V,
  Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019). “Welcome to the tidyverse.” _Journal of Open
  Source Software_, *4*(43), 1686. doi:10.21105/joss.01686 <https://doi.org/10.21105/joss.01686>.
  




